---
title: 'Lab 7: K-means Implementation'
author: "Lizzy Presland"
date: "March 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring and Preparing the Data

```{r}
# Q1
data_raw <- read.csv("snsdata.csv", stringsAsFactors = TRUE)
# Q2
summary(data_raw)
table(data_raw$gender)
```

#### Regarding Question 2: 

There is definitely missing values for gender - 2,724 missing values to be exact, or approximately 9% of the provided records. This is problematic because if these records were to be excluded, then almost 10% of the data is getting thrown out immediately. Additionally, the vast majority of the data is female, which means that the results will not evenly represent both female and male populations.

#### Question 3:

The `age` feature also has missing values. There are 5,086 examples with no age data, which corresponds to 16.95% of the dataset.

#### Question 4:

There are a few skewed data points in the `age` variable, with the maximum age provided being 106.927 years old. This is unrealistic, and those outlier values should be grouped into the `NA` category.

## Data Preparation - Handling Missing Values (`data$gender`)

#### Question 5:

Examples with missing values should ***NOT*** be excluded; this would cause 17%-25% of the dataset to be thrown out immediately.

#### Question 6:

```{r}
# add "no_gender" to gender factor levels
levels <- levels(data_raw$gender)
levels[length(levels) + 1] <- "no_gender"
# replace NA values with new factor
data_raw$gender <- factor(data_raw$gender, levels = levels)
data_raw$gender[is.na(data_raw$gender)] <- "no_gender"
# now to confirm this worked:
table(data_raw$gender)
```
