---
title: 'Lab 2: Linear Regression'
author: "Lizzy Presland"
date: "January 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Exploring and preparing the data

The data is taken from the .csv file posted on the class Canvas page.

#### Q1: Load the data

To load the data, we would normally use: 

```
ins_data <- read.csv("insurance.csv", header=TRUE)
```

As far as I can tell, we can load data with `stringsAsFactors=TRUE` in order for R to correctly parse the file and load a simple dataframe. However, we will want to use `stringsAsFactor=FALSE` in order to perform ***linear regression analysis*** on the data.

Therefore, our load function becomes:

```{r include=TRUE}
ins_data <- read.csv("insurance.csv", header=TRUE, stringsAsFactors = TRUE)
```

#### Q2: Data structure

```{r}
summary(ins_data)
```

#### Q3: Independent and dependent variables

The dependent feature is the **charges** column (the amount of money the insurance company can expect to pay for healthcare costs); the independent features are all others (**age**, **sex**, **bmi**, **number of children**, **tobacco use**, and **geographic region**).

#### Q4: Dependent feature analysis

It is best to analyze the dependent feature numerically; I shall use graphical and numerical methods.

#### Q5: Determine data's normal distribution

#### Histogram

```{r}
hist(ins_data$charges)
```

As the histogram shows, the data does not appear to be normally distributed; there appears to be a strong right skew to the data.

#### Numerical Methods

```{r include=FALSE}
cat("Mean: ")
mean(ins_data$charges)
cat("\nMedian: ")
median(ins_data$charges)
```

## Exploring relationships among features using a correlation matrix

```{r}
cor(ins_data$age, ins_data$charges)
cor(ins_data$bmi, ins_data$charges)
cor(ins_data$children, ins_data$charges)
```

**NOTE:** I was not able to get all 4 numerical values to generate a correlation matrix with the provided syntax. It would be helpful to review this in class.

## Visualizing relationships using a scatterplot matrix





## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
